# PDF & Word


### PDF

To create PDF documents, you'll need a $\TeX$ installation, and for that I suggest [tinytex](https://yihui.name/tinytex/). Once installed, you'll not need to use it directly, but will now be to knit documents to PDF.

With the Tex installation, you'll also be able to intermingle standard text, Latex, and R code seamlessly.


```
Here is some text!

\medskip

 ```{r}
x = rnorm(10)
 ```

$$\mathrm{SomeFormula} = Q_{123}^2$$

\pagebreak
```


### MS Word

As might be obvious, you'll need MS Word if you're going to create such documents. I wouldn't recommend this for anything, and so I will spend very little effort explaining to you how to do it.  

The main thing you need to know is how to use a template.  MS Word defaults have been terrible since its inception, so after customizing a document to look the way you want, you can then use it as a template.  Just specify the document you've created in the YAML as follows:


```{r word_yaml, eval=F}
output:
  word_document:
    reference_docx: my_template.docx
```


